---
title: "Lab9"
author: "Max Warriner"
output: word_document
---

Q1.

```{r}

library(lme4) # Load required libraries
library(car) 
library(dplyr)
library(rcompanion)
library(FSA)
library(emmeans)


strength = c(40,36,42,36,44,37,41,42,40,42,34,40,50,49,52,50,36,33,34,36,25,31,27,27,70,72,67,70,56,62,55,59,48,50,48,48)
pgm = c(rep("I",12),rep("II",12),rep("III",12))
dose = c(rep(c(rep("High",4),rep("Low",4),rep("Placebo",4) ),3))
camels = data.frame(pgm,dose,strength)
camels$pgm = as.factor(camels$pgm)
camels$dose = as.factor(camels$dose)
```

a.

1. Null: The training program of the camels does not affect their strength
Alt: The training program of the camels does affect their strength

2. Null: The Creatine dose of the camels does not affect their strength
Alt: The creatine dose of the camels does affect their strength

3. Null: The interaction between program and creatine does not affect their strength
Alt: The interaction between program and creatine does affect their strength


b. 

```{r}



boxplot(strength ~ interaction(pgm, dose), data = camels, 
        ylab = "Strength", xlab = "program", cex.lab = 1.2, cex.axis = 1.2)

# Filter the 'rat' dataset to create groups based on 'sex' and 'fat'
G1 = dplyr::filter(camels, pgm == "I" & dose == "High")$strength
G2 = dplyr::filter(camels, pgm == "II" & dose == "High")$strength
G3 = dplyr::filter(camels, pgm == "III" & dose == "High")$strength
G4 = dplyr::filter(camels, pgm == "I" & dose == "Low")$strength
G5 = dplyr::filter(camels, pgm == "II" & dose == "Low")$strength
G6 = dplyr::filter(camels, pgm == "III" & dose == "Low")$strength
G7 = dplyr::filter(camels, pgm == "I" & dose == "Placebo")$strength
G8 = dplyr::filter(camels, pgm == "II" & dose == "Placebo")$strength
G9 = dplyr::filter(camels, pgm == "III" & dose == "Placebo")$strength


testresult=scheirerRayHare(strength ~ pgm * dose, data = camels)
testresult


```

Interpretation:

After performing the Scheier-Ray hare Test, we observed that the distributions are similar so we can compare medians of the groups. After performing the test, we obtained significant results for the effect of program and dose on median strength, with p-values of 0.000063 and 0.026764 respectively. There was no significant interaction between the two. 

c. 

```{r}

dunnTest(camels$strength, camels$dose, method="bh")
dunnTest(camels$strength, camels$pgm, method="bh")

```

Interpretation:

We performed a follow up Dunn test on the effects of program and dose. When comparing dose level, we obtained significant results in the High-Placebo comparison (p-value = 0.0215383). When comparing program, we obtained significant results in the comparison between Programs 1 & 3 as well as 2 & 3, with p-values of 0.0005945692 and 0.0001689365 respectively. 

d. SUMMARY

We conducted a Scheier-Ray Hare test to determine if the dose level of creatine and the strength training program of camels affected their strength. After observing a similarity in distributions across all the 9 groups, we performed the test on the medians of the groups. After performing the test, we obtained significant results for the effect of dose level(p-value = 0.027) and program(p-value < 0.001) on the median of strength. The interaction was insignificant. A follow-up Dunn test was done and found a significant difference between the high and placebo doses for camels(p-value = 0.022) and between training programs 1&3(p-value < 0.001) and programs 2&3(p-value < 0.001). 


Q2.

```{r}
#Data preperation and creating a dataframe
cortex<-c(745,738,745,738,760,707,724,707,703,706,695,767,660,758,717,707,732,724,709,707)
food<-c(rep("Enriched",7),rep("Commercial",4),rep("Enriched",3),rep("Commercial",6))
environment<-c(rep("Stimulating",11),rep("Control",9))
rats<-data.frame(cortex, food, environment)
str(rats)
```

a.

```{r}
my.anova1a = lm(cortex ~ food * environment, data = rats)
my.anova1b = lm(cortex ~ environment * food, data = rats)

anova(my.anova1a) # PRINT ANOVA TABLE FOR THE FIRST MODEL
anova(my.anova1b) # PRINT ANOVA TABLE FOR THE SECOND MODEL

options(contrasts = c(unordered="contr.sum", ordered="contr.poly"))
my.anova1a = lm(cortex ~ food * environment, data = rats)
# Perform Type III ANOVA and print the ANOVA table for the first model
Anova(my.anova1a, type = 3) # For Type II: Anova(my.anova1a, type = 2)

```

b. Interpretation:

In both orders of doing type I SS, the food factor is significant (p-value < 0.05). However, when doing type III, no factor is significant. Because we have no specific reason for doing type I instead of type III, I would interpret this result as insignificant for all of the factors. 

Q3.

```{r}
#Data preparation
offspring = c(18,20,15,27,30,20,28,30,17,29,35,30,32,28,35,25,30,19,30,25,28,29, 32,38,29,35,39,30,40,38,28,36,29,30,37,33,39,42,47,38,51,42,48,39,55)
temp = c(rep("70",5), rep("75",5), rep("80",5),rep("70",5), rep("75",5), rep("80",5),
rep("70",5), rep("75",5), rep("80",5))
feeding = c(rep("1 Feeding",15), rep("2 Feedings", 15), rep("3 Feedings", 15))
#Create the  dataframe and assign the factors.
guppy = data.frame(offspring, temp, feeding)
guppy$temp = as.factor(guppy$temp)
guppy$feeding = as.factor(guppy$feeding)
str(guppy)
```

a.

```{r}
options(contrasts = c(unordered = "contr.sum", ordered = "contr.poly"))
my.anova = lmer(offspring ~ (1| temp) + (1 | feeding) + (1 | temp:feeding), data = guppy) # Fit the model
residuals = residuals(my.anova)

hist(residuals, freq=FALSE, xlab="Residuals", main="Residuals Histogram")
curve(dnorm(x, mean(residuals), sd(residuals)), add=TRUE, col='red')
qqnorm(residuals); qqline(residuals)
shapiro.test(residuals)

boxplot(offspring ~ interaction(feeding, temp), data = guppy, 
        ylab = "Offspring", xlab = "Feeding & Temp", cex.lab = 1.2, cex.axis = 1.2)


```

Interpretation:

Our analysis of the histrogram and qqplots show approximate normality. This is confirmed with a p-value of 0.5584 for the Shapiro normality test. When plotting the boxplots, we see an approximately similar distribution for each group, while we do have one outlier. 

b.

```{r,warning=FALSE}
model_no_temp = lmer(offspring ~ (1 | feeding) + (1 | feeding:temp), data = guppy, REML = FALSE)
anova(model_no_temp, my.anova)

model_no_feeding = lmer(offspring ~ (1 | temp) + (1 | feeding:temp), data = guppy, REML = FALSE)
anova(model_no_feeding, my.anova)

model_no_interact = lmer(offspring ~ (1 | feeding) + (1 | temp), data = guppy, REML = FALSE)
anova(model_no_interact, my.anova)

```

Interpretation:
1. Null: There is no significant variation in offspring due to water temperature in guppies
Alternative: There is significant variation in offspring due to water temp. 

2. Null: There is no significant variation in offspring due to feeding frequency.
Alternative: There is significant variation in offspring due to feeding frequency. 

3. Null: There is no significant variation in offspring due to the interaction between water temperature and feeding frequency.
Alternative: There is significant variation in offspring due to the interaction between water temperature and feeding frequency.

After performing the ANOVA test on all of the random factors, we obtain significant results for the effect of temperature and feeding frequency on variation with p-values of 0.006874a and 0.004664 respectively. The interaction was not significant.

c.

```{r}
anova_results = Anova(my.anova, type = "III") # Perform Type III ANOVA
variance_components = VarCorr(my.anova) # Extract variance components
residual_variance = attr(variance_components, "sc")^2 # Extract residual variance

# Calculate the variance for random effects
temp_variance = attr(variance_components$temp, "stddev")^2
feeding_variance = attr(variance_components$feeding, "stddev")^2
temp_feeding_variance = attr(variance_components$`temp:feeding`, "stddev")^2
# Calculate total variance (random effects and residual variance)
total_variance = temp_variance + feeding_variance + temp_feeding_variance + residual_variance

# Calculate percent variance contribution only if total variance is not zero
percent_variance = c("temp (Random)" = (temp_variance / total_variance) * 100,
    "feeding (Random)" = (feeding_variance / total_variance) * 100,
    "temp:feeding (Random)" = (temp_feeding_variance / total_variance) * 100,
    "Residuals" = (residual_variance / total_variance) * 100)
percent_variance

```

Interpretation:

After performing a follow up variance calculation, we found that 33% of the variation can be attributed to the water temperature, 43% of the variation can be due to the feeding frequency, and 0% of the variation is due to the interaction between the two factors. 

d.

```{r}

```

Interpretation:


e. SUMMARY

We performed a model two, two-way ANOVA to test if the feeding frequency and the water temperature contributed significant variation in the offspring of guppies. The data passed the normality assumption by passing the shapiro test for normality. We had one outlier in the data but we continued on with the ANOVA test. After performing the test, we obtained significant effects on the variation due to water temperature and feeding frequency with p-values of 0.006874 and 0.004664 respectively. After performing a follow up test on the variation, we found that 33% of the variation is due to the water temperature and 43% of the variaion is due to the feeding frequency. 

Q4. 

a.

```{r}

1-pf(2.98,19,6)

```

Interpretation:
The null hypothesis is that the elevation does not add significant variation to the biodiversity in soil invertebrates. They cannot reject the null hypothesis, because their p-value with a test statistic F(19&6-df) is 0.09, greater than 0.05.

b.

```{r}



```

Interpretation:

Extra Practice 1.

a.

b.

c.

d.

e.

f.

g.